Liveness-Based Garbage Collection for Lazy Languages

### 2015-04-28 paper/paper.tex

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.

In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

@@

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\psi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\psi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

\section{Implementation}
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.  An important difference in  GC of lazy language compared to
an  eager language is  the point  of garbage  collection.  In  case of
eager  languages only  \CONS~ points  and  function calls  are the  GC
points. In  a lazy  language every \LET~  statement is a  potential GC
point. Hence we need to store liveness automata corresponding to every
\LET~  statement in the  program. This  leads to  an explosion  in the
number of  automata. \comment{Directly say that we share the automata}From our  observation that along an  {\em E-path}
all program points have the same liveness, we can reduce the number of
automata state required  by sharing the same automata  for all program
points along an {\em E-path}. In  the example program, we only need to
create automata  for program  points \ldots, since  they are  the only
distinct {\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

>>>>

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

@@

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\psi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\psi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\psi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================
In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
 and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

@@

\section{Implementation}
% GC assumptions
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.  An important difference in  GC of lazy language compared to
an  eager language is  the point  of garbage  collection.  In  case of
eager  languages only  \CONS~ points  and  function calls  are the  GC
points. In  a lazy  language every \LET~  statement is a  potential GC
point. Hence we need to store liveness automata corresponding to every
\LET~  statement in the  program. This  leads to  an explosion  in the
number of  automata. \comment{Directly say that we share the automata}From our  observation that along an  {\em E-path}
all program points have the same liveness, we can reduce the number of
automata state required  by sharing the same automata  for all program
points along an {\em E-path}. In  the example program, we only need to
create automata  for program  points \ldots, since  they are  the only
distinct {\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}
